---
title: 'Homework 6: Land Suitability Analysis'
author: "Alfredo Rojas"
date: "3/27/2020"
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Land suitability analysis for Durham County. 

Objective: Deciding where a landfill should go.

Since landfilles are unwanted forms of land use, figuring out where one should be placed is often contentious. In this imagined example, I will build on Nikhil Kaza's [tutorial](http://sia.planning.unc.edu/post/2020-03-06-land-suitability-analysis/land-suitabilty-with-ahp-wlc/) on land suitability analysis. This assignment builds on his analysis by including indicators of environmental justice and other environmental criteria. These criteria are:

(1) Do not disproportionately expose minority communities to environmental hazards
(2) Avoid putting hazardous sites near protected areas
(3) Looking at soil types best suitable for landfills


I learned how to get ACS income data and visualizing it with the help of Kyle Walker's [tutorial](https://walkerke.github.io/tidycensus/articles/spatial-data.html).

```{r, results='hide', warning = FALSE, message=FALSE}

library(raster)
library(rasterVis)
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)
library(fasterize)
library(sf)
library(tmap)
library(tidycensus)
library(RColorBrewer)

# From Nikhil Kaza's blog, he uses this template raster as a basis to analyze other datasets. 
lulc_data = raster("land_suit/c11_37063.img")
template_raster = reclassify(lulc_data, cbind(0, 100, 0), right=FALSE)
project_crs = crs(lulc_data)

# explore tidycensus variables
# v18 = load_variables(2018, data = "acs5", cache = TRUE)

# bring other data for suitability analsyis
durham_tract = tidycensus::get_acs(geography = "tract", 
                        variables = "B19013_001", 
                        #year = "2018", 
                        state = "NC", 
                        county = "Durham", 
                        geometry = T)

durham_tract = st_transform(durham_tract, crs=project_crs) 

# Look at scaled income across Durham county tracts
durham_tract %>% 
  mutate(scale_val = scale(estimate, center = F)) %>%
  ggplot(aes(fill = scale_val)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c()

```

Look at race and ethnicity distribution of Durham county. Because I can get this data from the decennial census, I can visualize this by census blocks. 

```{r, results='hide', warning = FALSE, message=FALSE}

# get variables for race and ethnicity population counts
race_code <- c(White = "P005003", 
              Black = "P005004", 
              Asian = "P005006", 
              Hispanic = "P004003")

durham_blocks = tidycensus::get_decennial(geography = "block", 
                        variables = race_code, 
                        year = "2010", 
                        state = "NC", 
                        county = "Durham", 
                        geometry = TRUE,
                        summary_var = "P001001")

durham_blocks = st_transform(durham_blocks, crs = project_crs)

durham_blocks %>% 
  mutate(pop_pct = value / summary_value) %>%
  ggplot(aes(fill = pop_pct)) +
  facet_wrap(~variable) +
  geom_sf(color = NA) +
  scale_fill_viridis_c()


```

Perhaps one angle we can take is to ensure that minority communities are not put at disproportionate risk of exposure to environmental contaminants. Racial disparities in exposure to environmental risks is termed [environmental racism](https://www.theatlantic.com/politics/archive/2018/02/the-trump-administration-finds-that-environmental-racism-is-real/554315/) and has been evidenced in cases of lead poisoning and particulate matter in the air affecting minority communities more than wealthier predominantly white communities. See also this [interactive article](http://www.msnbc.com/interactives/geography-of-poverty/se.html) on poverty and environmental injustice in the the American South.

## Protected Areas Database

```{r}
# read in protected are designation
# pad_data = st_read("pad_nc/PADUS2_0Designation_NC.shp") %>%
#   st_as_sf()
# 
# p1 = tm_shape(pad_data) +
#   tm_polygons("Mang_Type")
# 
# project and clip to Durham
# pad_clip = pad_data %>%
#   st_transform(project_crs) %>%
#   st_crop(st_bbox(lulc_data))
# 
# st_write(pad_clip, paste0(getwd(), "/", "land_suit", "/", "pad_durham.shp"))
```

Clip PAD data to Durham County. Just to visuzlize where these protected areas are spatially, I'll plot it against the Durham tract-level data. 

```{r}

# using subset of data created for this specific purpose (code commented out above)
pad_clip = st_read("land_suit/pad_durham.shp")

p1 = tmap::tm_shape(durham_tract) +
  tm_polygons(col = NA)

p2 = tmap::tm_shape(pad_clip) +
  tm_polygons("Mang_Type") 

p1 + p2

```

One criteria that we can strive for is that the landfill be sufficiently far away from protected areas. Following, Nikhil Kaza's [blog](http://sia.planning.unc.edu/post/2020-03-06-land-suitability-analysis/land-suitabilty-with-ahp-wlc/), we can convert the polygon data into raster, and determine different distances from locations of interest using the `gridDistance()` function. 

```{r}

pad_ras = pad_clip %>%
  fasterize(raster = template_raster) %>%
  gridDistance(origin = 1) %>%
  mask(lulc_data) %>% # Creates new raster that has the same values as lulc_data, except for the cells that are NA (or other maskvalue) in a 'mask'.
  reclassify(rcl = matrix(c(0,500,1, 500,1000,2, 1000,2000,3, 2000,4000,4, 4000,Inf,5), ncol=3, byrow = T), include.lowest=T)

pad_ras %>% ratify %>% levelplot(att="ID", main = "Distance from PAD", col.regions=brewer.pal(6, "PuBu"))

```

## Soils

Another domain land suitability analysis has a role is looking at the kinds of soils in an area to see what soil type is best suited for the specific land use. We are able to achieve this by using [soil data](https://nrcs.app.box.com/v/soils) from the National Resources Conservation Service.

```{r}
# nc_soil = st_read("soils/wss_gsmsoil_NC_[2016-10-13]\\spatial\\gsmsoilmu_a_nc.shp") %>% 
#   st_as_sf() %>%
#   st_transform(project_crs) %>%
#   st_crop(st_bbox(lulc_data))
# 
# st_write(nc_soil, paste0("land_suit", "/", "durham_soil.shp"))

dur_soil = st_read("land_suit/durham_soil.shp")

tmap::tm_shape(dur_soil) +
  tm_polygons("MUSYM")

```

The MUSYM labels have to changed to identify soil types. We can do that with the below code, using one of the .txt documents that came with the original dataset:

```{r}

# create lookup table for durham
# soil_codes = unique(dur_soil$MUSYM)
# lookup_table = read_delim("soils/wss_gsmsoil_NC_[2016-10-13]\\tabular\\mapunit.txt", 
#                           col_names = FALSE, delim = "|", skip_empty_rows = TRUE) %>% 
#   dplyr::select(X1, X2) %>%
#   filter(X1 %in% as.character(soil_codes))

# save for later:
# write_csv(lookup_table, "land_suit/lookup_table.csv")
#test
lookup_table = read_csv("land_suit/lookup_table.csv")

# creates vectors from tibble
soil_names = pull(lookup_table[,2]) %>%
  str_replace(" \\(.*\\)", "")
code_lookup = pull(lookup_table[,1])

# add labels
count = 1
dur_soil$labels = NA
for(i in code_lookup){
  dur_soil[dur_soil$MUSYM == i, ]$labels = soil_names[count]
  count = count + 1
}

# plot with labels
tmap::tm_shape(dur_soil) +
  tm_polygons("labels") +
  tm_layout(legend.outside = T)

```

Now that we have labeled soil data, we can perform similar raster operations on this dataset in order to incorporate this into similar raster operations we've done.

```{r}
soil_ras = dur_soil %>%
  fasterize(raster = template_raster, by = "labels")
  

plot(soil_ras)

```

With this, we are now able to identify which soil levels are most suitable for a land fill and select that raster layer for further land suitability analysis. 


