---
title: 'Assignment 1: Tree Census in NY, 2005 - 2015'
author: "Alfredo Rojas"
date: "1/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Explore Tree Census data for 2005

```{r library results = "hide", message = FALSE}

library(tidyverse)
library(sf)
path_to_file = file.path("..", "data", "2005_Street_Tree_Census.csv")
tree_2005 = path_to_file %>% read_csv()
```

## Explore data

Clean up the data using `group_by()` and `summarise()`. 

```{r clean}
head(tree_2005)

byBoro = tree_2005 %>%
  count(nta_name, boroname, spc_common, status, sort = TRUE)

head(byBoro, n = 10)
  
```

Let's visualize

```{r visual}

byBoro %>% ggplot() +
  geom_bar(aes(x = boroname, fill = status), position = "dodge") +
  coord_flip()
```

More visuals

```{r visual2}

treeLocs = tree_2005 %>%
  select(OBJECTID, spc_common, nta_name, boroname, latitude, longitude)



```


### notes

-Distribution of Income by borough. . . 

Positives: Narrative, single-variable representation of tree density

Negatives: Assumed spatial knowledge of NY, forces the comparison between borough (which by nautre means that income is segmented by boroughts), unclear whether red/green gradation of borough map is is within borough or total area. . . No legend determining intervals. . . 

To-do: 
(1) do a graph of distribution of income by borough. . . 
(2) do a map based on neighborhood. . . 

--Finding income data
--Finding neighborhood data

Tasks:
--Normalize by greenspace
--