---
title: "Final Paper Exploratory Data Analysis"
author: "Alfredo Rojas"
date: "3/10/2020"
output:
  github_document
---


## Final Project for PLAN 672: Initial Summary

This project will explore the relationship between food insecurity indicators in West Africa and relevant environmental and household-related variables. This preliminary paper will rely on DHS survey data. For this specific analysis, the survey data is restricted to 2011 since the DHS standardizes all their variables across survey years, i.e. the variables all have the similar code names for easy comparison. 

At the outset, vaiables related to food security, according to Gubert, et al. (2010) are: per capital income, years of schooling, race and gender of HH head, urban/rural residence, access to public water supply, presence of children, total number of household inhabitants, and state of residence. 

This particular project will begin exploring some of these variables within the DHS data. Later, I will try to incorporte othr related variables, such as land-use/land-cover change, NDVI, and/or climate data. As I explore the literature in more detail, I will determine what will be a feasible route to take for this final project. 

## What is an HAZ score

Height for weight z-scores, or standard deviation scores, are indices to measure linear growth potential in children, where lower scores indicate stunting as a result of poor health conditions and undernutrition (see, [WHO](https://www.who.int/nutgrowthdb/about/introduction/en/index2.html) article).




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, include = FALSE)
```

### Ivory Coast Data, 2011

Let's explore some of the variables in the data and see what may be of interest. Later, I will bring up what variables are commonly used in the literature and whether they are present in this particular dataset. The first thing I wanted to look for were differences between rural and urban households. This distinction will be important to see what kinds of variables impact food security indicators. For example, does living in rural or urban settings impact food security measures? If so, to what extent? These are some of the questions I begin asking as I explore the DHS data to see what variables I'm able to work with. 

The `haven` package allows R to read and interpret data stored in .DTA format. In this format, categorical variables are 
stored as `haven_labelled` class, meaning the categorical variables used in observations correspond to label names. You can insert these label names using the `haven::as_factor()` function, as seen in the `ggplot` code chunk. 

Independent variables to use: HHID, hv001 (cluster number), hv012 (num of de jure HH members), hv025 (urban/rural), HV107? (years of highest education--maybe), hv209 (has fridge), hv244 (owns land), hv245 (ag area), hv270 ("wealth index").

Dependent variables wil lbe: hc70_x ("height-for-age SD WHO"), 

```{r read, messages = FALSE}

library(haven)
library(tidyverse)
library(foreign)
library(sf)
library(dplyr)
library(plyr)
library(ggplot2)
# library(ipumsr)

# vars_that_should_be_doubles <- c("NDVI", "CROPLAND", "PRECIP", "TEMPMAX")
# 
# ddi <- read_ipums_ddi("ipums_data/idhs_00007.xml")
# ddi$var_info <- ddi$var_info %>% 
#   mutate(
#     var_type = if_else(
#       var_name %in% vars_that_should_be_doubles,
#       "double",
#       var_type)
#   )
# data <- read_ipums_micro(ddi)


data <- read_dta("ipums_data/idhs_00007.dta")
data_bf <- read_dta("ipums_data/ipums_bf2010.dta")

```

Removing missing values. . . Alderman and Heady (2018) drop observations with HAZ scores below -6 and above 6. . . 

Note; Cropland has some missing variables. . . have to recode some. . . 

```{r clean}
data_small <- data %>%
  filter(
    hwchazwho >= -600,
    hwchazwho <= 600,
    hwcwhzwho >= -600,
    hwcwhzwho <= 600
  ) %>%
  select(-c(hwcwazwho, hwcwhzwho))

# create vector of relevant column names then iterate through dataframe to change -998 to NAs.
ndvi_names <- data[grep("ndvi_", names(data))] %>% names()
for(i in ndvi_names){
    data_small[c(i)] <- na_if(data_small[c(i)], -998) # na_if takes vector
}

# chagne temp -998 to NAs
temp_names <- data[grep("tempmax_", names(data))] %>% names()
for(i in temp_names){
  data_small[c(i)] <- na_if(data_small[c(i)], -998)
}

# remove precip -998 and replace with NAs
precip_names <- data[grep("precip_", names(data))] %>% names()
for(i in precip_names){
  data_small[c(i)] <- na_if(data_small[c(i)], -998)
}

# change crop -998 to NAs
data_small$cropland <- na_if(data_small$cropland, -998)

# keep complete observations, remove labels that `haven` preserves from .DTA file for ML model
data_small <- data_small[complete.cases(data_small), ] %>% zap_labels() 

# final data for ML model. . . 
data_final <- data_small %>%
  as.data.frame() %>%
  mutate(
    ndvi_mean = rowMeans(data_small[ndvi_names]),
    temp_mean = rowMeans(data_small[temp_names]),
    precip_mean = rowMeans(data_small[precip_names]),
    stunting = if_else((hwchazwho/100) < -2, 1, 0)
  ) %>% 
  select(-c("ndvi", "precip", "tempmax", "hhresident", ndvi_names, temp_names, precip_names)) %>%
  select(-c(1:5, 7:8, 10, 11, 12))

```

## Looking at the distribution of height for age scores

```{r haz-dens}

group_med = ddply(data_small, "urbanhh", summarise, hh.med = median(hwchazwho/100))
group_med$urbanhh <- as.factor(group_med$urbanhh)

data_small$urbanhh <- as.factor(data_small$urbanhh)

ggplot(data_small, mapping = aes(x = hwchazwho/100, fill = urbanhh, group = urbanhh)) + 
  geom_density(size = 0.5, alpha = 1/3) +
  labs(
    title = "Density of Height-for-Age-Z-Scores (HAZ) in Urban and Rural Areas, Côte d'Ivoire",
    x = "Density",
    y = "HAZ Score"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = group_med, 
             mapping = aes(xintercept = hh.med, color = urbanhh),
             show.legend = F, linetype = "dashed")

```

## Looking at the distribution for proportion of crop land around clusters

```{r crop-dens}

crop_land_mean <- ddply(data_small, "urbanhh", summarise, cl_mean = mean(cropland))
crop_land_mean$urbanhh <- as.factor(crop_land_mean$urbanhh)
data_small %>%
  ggplot(mapping = aes(x = cropland, fill = urbanhh, group = urbanhh)) +
  geom_density(alpha = 1/3) +
  labs(
    title = "Density of cropland in Urban and Rural Areas",
    subtitle = "Proportion of cropland within 10-km buffer of cluster",
    x = "Density",
    y = "Proportion of cropland"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = crop_land_mean, 
             mapping = aes(xintercept = cl_mean, color = urbanhh),
             show.legend = F)
  
```

## Scatter plot between cropland and HAZ

```{r crop-haz}
data_small %>%
  ggplot(mapping = aes(x = scale(cropland), y = scale(hwchazwho/100))) +
  geom_point() +
  geom_abline()

data_final %>%
  filter(hwchazwho/100 < 0) %>%
  ggplot(mapping = aes(x = scale(temp_mean), y = scale(hwchazwho/100))) +
  geom_point() +
  geom_abline()

data_final %>%
  ggplot(aes(dhsid)) +
  geom_col()

```


## Exploring NDVI and temperature values

So, I'm thinking, for NDVI, since they have 60 months BEFORE survey data, I can take the first 12 months, calculate the mean annual total NDVI (MATN) and then take the 12 months from the survey date to do the same thing. Then I can subtract T2 from T1, which can be my variable. . . Maybe I can do this for all environmental variables. . . SO these variables are representing changes from 2006 to 2011. Or maybe I can just take the average for all months.

Maybe I have to add min. . . 

Plotting for one CLUSTER (WHERE? -- maybe show three CLUSTERS, South, Central, North), shows the cyclic nature of NDVI over time. I am able to identify the month, but I have to re-download the data again. 

```{r ndvi-temp}

data_final %>%
  ggplot(mapping = aes(x = precip_mean, y = ndvi_mean)) +
  geom_point() +
  geom_smooth()

ndvi_series <- data_small %>%
  select(dhsid, ndvi_names) %>%
  group_by(dhsid) %>%
  summarise_all(first)

# plotting an individual cluster over time. . . 
ndvi_series[ndvi_series$dhsid == "CI201200000301", ] %>%
  pivot_longer(-c(dhsid, dhsid), names_to = "month", values_to = "ndvi") %>%
  ggplot(mapping = aes(x = month, y = ndvi, group = 1)) +
  geom_point() + 
  geom_line() 

```

```{r ndvi-total}

ndvi_total <- data %>%
  select(dhsid, ndvi_names) %>%
  left_join(regions, by = c("dhsid" = "DHSID")) %>%
  st_as_sf() %>%
  st_drop_geometry() %>% # drop geom only works with `sf` objects
  group_by(dhsid) %>%
  summarise_all(first) # first is a function first()
 
ndvi_total[ndvi_total$DHSREGNA == "Ville D'Abidjan", ][1, ] %>%
  pivot_longer(-c(dhsid, DHSREGNA), names_to = "month", values_to = "ndvi") %>%
  ggplot(mapping = aes(x = month, y = ndvi, group = 1)) +
  geom_point() + 
  geom_line()

ndvi_total[ndvi_total$DHSREGNA == "Nord-Ouest", ][6, ] %>%
  pivot_longer(-c(dhsid, DHSREGNA), names_to = "month", values_to = "ndvi") %>%
  ggplot(mapping = aes(x = month, y = ndvi, group = 1)) +
  geom_point() + 
  geom_line()

```


```{r mlTest, include = TRUE}
library(caret)

data_final2 <- rbind(data_final, data_final_bf)

data_final2$urbanhh <- as.factor(data_final2$urbanhh)
# test2$hhresident <- as.factor(test2$hhresident)  
data_final2$aglandyn <- as.factor(data_final2$aglandyn)
data_final2$wealthqhh <- as.factor(data_final2$wealthqhh)
data_final2$stunting <- as.factor(data_final2$stunting)

# saveRDS(subset_data, "test.rds") # create test data 

data_ready <- data_final2  %>% 
  zap_label() %>%
  select(-c("hhid", "dhsid", "hwchazwho"))
rm(data_final2)

# create index to get training samples
train_index <- createDataPartition(data_ready$stunting, p=0.80, list=FALSE)
# get test subset
test_dataset <- data_ready[-train_index,]
# this will be used for training subset
train_dataset <- data_ready[train_index,]

set.seed(12)
control <- trainControl(method="repeatedcv", repeats =3, 
                        classProbs= F, summaryFunction = multiClassSummary) 
# classProbs = F avoids the error of syntactically valid names. . . I don't get it though. . . 
# control <- trainControl(method="cv", number = 10)
rf_haz <- train(stunting ~ ., data = train_dataset, method = "rf", 
                trControl = control, preProcess = c("center", "scale"))

haz_predict <- predict(rf_haz, test_dataset)
confusionMatrix(factor(haz_predict), factor(test_dataset$stunting)) # gives error: the data cannot have more levels than the reference

# I don't think Confusion matrix is appropriate for continuous vars
# table(haz_predict)
# table(test_dataset$hwchazwho)
# 
# u <- union(haz_predict, test_dataset$hwchazwho)
# t <- table(factor(haz_predict, u), factor(test_dataset$hwchazwho, u))
# confusionMatrix(t)

```

## Exploratory Portion

I am going to create a subset data frame to explore some immediate variables of interest. Note the use of `as_factor()` in the `ggplot()` function. 

```{r}

# compare urban/rural numbers, 1 = urbn, 2 = rural
p1 <- subset_hh %>%
  group_by(urban_rural) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = as_factor(urban_rural), y = n)) + # as_factor a `haven` function, uses labels for categorical responses
  geom_bar(stat = "identity")

p1
# ggsave("rural_urban_bar.png", dpi = 300)

```

I can start looking at the distribution of certain variables of interest. Let me start with number of household members per each household observed. 

```{r}

p2 <- ggplot(subset_hh, aes(num_hh_mem)) +
  geom_density()

p2
# ggsave("hh_density.png", dpi = 300)
```

Now, I'll look at agricultural area in hectares for each household that has a farm. Note that DHS codes Unknown or Missing data as 998 and 999, respectively, so we will have to handle this. The problem is that when we remove all responses that were 998, I lose around 4,000 observations, which is a lot. One solution around this is to imputate the data for all these values once I decide which model to run.

```{r}
d2 <- subset_hh %>%
  filter(is.na(ag_area) != TRUE, ag_area < 998) %>%
  select(ag_area)

p3 <- ggplot(d2, aes(ag_area)) +
  geom_density()
p3

# ggsave("ag_dens.png", dpi = 300)
```

Also, I can look at the relationship between the two variables. 

```{r}

# plot houehold member number by agri area
p4 <- subset_hh %>%
  ggplot(aes(x = num_hh_mem, y = ag_area)) +
  geom_point() +
  geom_smooth(method = "loess")

p4

# ggsave("ag_hh_compare.png", dpi = 300)
```

The DHS has another variable that may be of interest: Wealth Index. This index apparently measures the relative wealth of households and categorizes them into quintiles. 

```{r}
# plot wealth vs. agricultural area
 p5 <- subset_hh %>%
  ggplot(aes(x = wealth_index, y = ag_area)) +
  geom_col(aes(fill = as_factor(wealth_index)))

# ggsave("ag_wealth.png", dpi = 300)
```

One thing that may be of interest is looking at the distribution of agricultural area by wealth category to see variation within a variable. The spread of the data is interesting. The different catgories seem to follow a similar pattern but in different magnitudes. 

```{r}

# compare distribution of agricultural area across wealth indices
p6 <- subset_hh %>%
  ggplot(aes(x = ag_area, colour = as_factor(wealth_index))) +
  geom_freqpoly(binwidth = 50)

# ggsave("ag_wealth_density.png", dpi = 300)
```


### Geospatial

Exploring the shapefile that comes with the 2011-12 data for Côte d'Ivoire.

```{r}
library(sf)
geo_path <- file.path("CI_Data", "geo-data-2011", "CIGE61FL", "CIGE61FL.shp")
shp_2011 <- st_read(geo_path)

plot(st_geometry(shp_2011)) 

head(shp_2011)

regions <- shp_2011 %>%
  select(DHSID, DHSREGNA)


```

The shapefile has the field `DHSCLUST` which corresponds to the `hv001` which is the cluster number in the hosuehold-level dataset. 

