---
title: "Final Paper"
author: "Alfredo Rojas"
date: "4/27/2020"
output:
  github_document
---

## Final Project for PLAN 672: Initial Summary

This preliminary project will work towards building a predictive model for food insecurity indicators in West Africa, using relevant environmental and household-related variables, as well as relying on a machine learning approach. This preliminary paper will rely on [DHS-IPUMS](https://www.idhsdata.org/idhs/) survey and environmental data for Burkina Faso and Ivory Coast (Boyle, King, and Sobek 2019). For this specific analysis, the survey data is restricted to 2010 and 2011. Since the DHS standardizes all their variables across survey years, i.e. the variables all have similar code names for easy comparison. 

One way to measure food insecurity in a region is to look at its consequences for health. For example, one method to measure these consequences is by looking at malnutrition in children, which leads to growth faltering, a popular overall indicator of a child's health, as well as an indicator of social inequalities (de Onis and Branca 2016). Growth faltering is measured in height-for-age z-score (HAZ), where scores below -2 mean stunted growth. In West and Central Africa, nearly 32% of children are stunted (UNICEF 2020). Measuring stunting, therefore, is a crucial avenue to study the drivers of food insecurity.

In this project, I rely on survey data collected through the Demographic Health Survey (DHS) vaiables related to food security, as well as contextual variables relating to environmental conditions from IPUMS. Common variables, according to Gubert, et al. (2010) are: per capital income, years of schooling, race and gender of HH head, urban/rural residence, access to public water supply, presence of children, total number of household inhabitants, and state of residence. In the DHS-IPUMS data, I have access to and will use:

* Number of household members
* Urban/rural household residence
* Agricultural holdings
* Wealth index
* HAZ 
* Proportion of cropland within household cluster
* mean monthly maximum Normalized Difference Vegetation Index (NDVI) across, 60 months before and 11 months after survey date
* mean monthly maximum precipitation across, 60 months before and 11 months after survey date
* mean monthly maximum temperature across, 60 months before and 11 months after survey date

The dependent variable will be HAZ score, while the rest will act as explanatory variables. 

## What is a HAZ?

Height for age z-scores, or standard deviation scores, are indices to measure linear growth potential in children under age 5, where lower scores (HAZ < -2) indicate stunting as a result of poor health conditions and undernutrition (see, [WHO](https://www.who.int/nutgrowthdb/about/introduction/en/index2.html) article). While HAZ is measured for children at any age under 5 years old, recent research as argued that HAZ scores for children between 0-23 months should be analyzed separately since the impacts of malnutrition or any protective factors have not manifested until after 23 months (Alderman and Headey 2018). For the sake of simplicity in this analsis, I do not differentiate between age. But as I develop this project, this separation will be necessary.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, echo = FALSE)
```

## Burkina Faso (2010) and Ivory Coast (2011) data

I chose to compare the 2010 and 2011 data for Burkina Faso and Ivory Coast (Côte d'Ivoire in French), respectively. While it should not be underestimated that regional differences exist between the tropical, Sudanian savanna, and Sahelian regions going south to north, I do not explicitly account for these differences in this preliminary stage. 

```{r read, messages = FALSE}

library(haven)
library(tidyverse)
library(foreign)
library(sf)
library(leaflet)
library(dplyr)
library(plyr)
library(ggplot2)
library(caret)

data <- read_dta("ipums_data/idhs_00007.dta")
data_bf <- read_dta("ipums_data/ipums_bf2010.dta")

```

Below, I show all of the GPS points of the sample clusters for these two surveys. GPS points are randomly offset by approximately 5 to 10 kilometers for rural households to ensure anonymity. Note that the data includes one erroneous point (perhaps a displaced point) that appears in the ocean! 

```{r, messages = FALSE, results = 'hide'}
ci_geo_path <- file.path("CI_Data", "geo-data-2011", "CIGE61FL", "CIGE61FL.shp")
bf_geo_path <- file.path("BF_Data", "geo-data-2010", "BFGE61FL", "BFGE61FL.shp")
ci_gps <- st_read(ci_geo_path) %>% st_transform(4326)
bf_gps <- st_read(bf_geo_path) %>% st_transform(4326)

leaflet() %>% 
  addCircles(data = ci_gps) %>%
  addCircles(data = bf_gps) %>%
  addTiles()
```

## Methods

This analysis uses the R statistical programming language to clean, process, visualize, and analyze the DHS-IPUMS data. The predictive model will rely on a Random Forest binary classifier, testing the importance of each explanatory variable in predicting whether a child has an HAZ below -2 or not. I rely on the `caret` package to construct the model. By nature of constructing a binary classifier, I categorized HAZ scores below -2 as "1" and anything above as "0". All HAZ values above 6 and below -6 are dropped.

The environmental variables, like maximum monthly NDVI, precipitation, and temperature are provided by IPUMS for 60 months before the survey month and 11 months after at the cluster level. This analysis takes the total average for these variables. For individual HAZ values, there are a total of 10,164 observations.

```{r clean}

clean_ipums <- function(x) {
  #'
  #' This function cleans the IPUMS data to extract a binary categorization of stunted and not-stunted
  #' among the other relevant variables for this projects's analysis
  #' 
  
  data_small <- x %>%
  filter(
    hwchazwho >= -600,
    hwchazwho <= 600,
    hwcwhzwho >= -600,
    hwcwhzwho <= 600
  ) %>%
  select(-c(hwcwazwho, hwcwhzwho)) # decided not to use these
  
  # create vector of relevant column names then iterate through dataframe to change -998 to NAs.
  ndvi_names <- x[grep("ndvi_", names(x))] %>% names()
  for(i in ndvi_names){
      data_small[c(i)] <- na_if(data_small[c(i)], -998) # na_if takes vector
  }
  
  # chagne temp -998 to NAs
  temp_names <- x[grep("tempmax_", names(x))] %>% names()
  for(i in temp_names){
    data_small[c(i)] <- na_if(data_small[c(i)], -998)
  }
  
  # remove precip -998 and replace with NAs
  precip_names <- x[grep("precip_", names(x))] %>% names()
  for(i in precip_names){
    data_small[c(i)] <- na_if(data_small[c(i)], -998)
  }
  
  # change crop -998 to NAs
  data_small$cropland <- na_if(data_small$cropland, -998)
  
  # keep complete observations, remove labels that `haven` preserves from .DTA file for ML model
  data_small <- data_small[complete.cases(data_small), ]
  
  # create this for ML model 
  data_final <- data_small %>%
    as.data.frame() %>%
    mutate(
      ndvi_mean = rowMeans(data_small[ndvi_names]),
      temp_mean = rowMeans(data_small[temp_names]),
      precip_mean = rowMeans(data_small[precip_names]),
      stunting = if_else((hwchazwho/100) < -2, 1, 0) # create binary variable for RF classifier
    ) %>%
    select(-c("ndvi", "precip", "tempmax", # remove for final model
            "hhresident", ndvi_names, temp_names, precip_names)) 
  
  return(data_final)
}

ci_clean <- clean_ipums(data) %>%
  select(-c(1:12)) # get rid of other unnecssary vars for analysis

bf_clean <- clean_ipums(data_bf) %>%
  select(-c(1:11))

```

## Looking at the distribution of height for age scores

In general, rural HAZ values are lower than urban HAZ for both Ivory Coast and Burkina Faso. Overall, median HAZ are below zero. In the charts below, I provide probability density graphs to examine the general distribution of scores across rural and urban areas.

```{r haz-dens}

# Density plot for CI
group_med = ddply(ci_clean, "urbanhh", summarise, hh.med = median(hwchazwho/100))
group_med$urbanhh <- as.factor(group_med$urbanhh)

ci_clean$urbanhh <- as.factor(ci_clean$urbanhh)

# plot HAZ density
ggplot(ci_clean, mapping = aes(x = hwchazwho/100, fill = urbanhh, group = urbanhh)) + 
  geom_density(size = 0.5, alpha = 1/3) +
  labs(
    title = "Density of Height-for-Age-Z-Scores (HAZ) in Urban and Rural Areas, \nCôte d'Ivoire (2011)",
    x = "HAZ",
    y = "Density"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = group_med, 
             mapping = aes(xintercept = hh.med, color = urbanhh),
             show.legend = F, linetype = "dashed")

```


```{r bf-dens}

# look at HAZ density plot for Burkina
group_med2 = ddply(bf_clean, "urbanhh", summarise, hh.med = median(hwchazwho/100))
group_med2$urbanhh <- as.factor(group_med2$urbanhh)

bf_clean$urbanhh <- as.factor(bf_clean$urbanhh)

# plot HAZ density for BF
ggplot(bf_clean, mapping = aes(x = hwchazwho/100, fill = urbanhh, group = urbanhh)) + 
  geom_density(size = 0.5, alpha = 1/3) +
  labs(
    title = "Density of Height-for-Age-Z-Scores (HAZ) in Urban and Rural Areas, \nBurkina Faso (2010)",
    x = "HAZ",
    y = "Density"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = group_med2, 
             mapping = aes(xintercept = hh.med, color = urbanhh),
             show.legend = F, linetype = "dashed")
```

## Looking at the distribution for proportion of crop land around clusters

Another consideration is to explore the amount of cropland that exists around survey sites across urban and rural areas. As would be expected, the mean proportion of cropland is lower in urban areas than in rural areas. However, the rural and urban distributions are very different for Ivory Coast than for Burkina Faso. Ivory Coast's distribution for rural areas seems to gravitate around the mean, whereas in the urban distribution you find more distinct peaks. In Burkina Faso's distribution, both urban and rural seem to follow a similar pattern. 

```{r crop-dens}

# plot crop density for Cote d'Ivoire
crop_land_mean <- ddply(ci_clean, "urbanhh", summarise, cl_mean = mean(cropland))
crop_land_mean$urbanhh <- as.factor(crop_land_mean$urbanhh)
ci_clean %>%
  ggplot(mapping = aes(x = cropland, fill = urbanhh, group = urbanhh)) +
  geom_density(alpha = 1/3) +
  labs(
    title = "Density of cropland in Urban and Rural Areas, Côte d'Ivoire (2011)",
    subtitle = "Proportion of cropland within 10-km buffer of cluster",
    x = "Density",
    y = "Proportion of cropland"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = crop_land_mean, 
             mapping = aes(xintercept = cl_mean, color = urbanhh),
             show.legend = F, linetype = "dashed")
  
```

```{r bf-crop-dens}

# plot crop density for Burkina
crop_land_mean2 <- ddply(bf_clean, "urbanhh", summarise, cl_mean = mean(cropland))
crop_land_mean2$urbanhh <- as.factor(crop_land_mean2$urbanhh)
bf_clean %>%
  ggplot(mapping = aes(x = cropland, fill = urbanhh, group = urbanhh)) +
  geom_density(alpha = 1/3) +
  labs(
    title = "Density of cropland in Urban and Rural Areas, Burkina Faso (2010)",
    subtitle = "Proportion of cropland within 10-km buffer of cluster",
    x = "Density",
    y = "Proportion of cropland"
  ) +
  scale_fill_discrete(name = "Residence", labels = c("Urban", "Rural")) +
  geom_vline(data = crop_land_mean2, 
             mapping = aes(xintercept = cl_mean, color = urbanhh),
             show.legend = F, linetype = "dashed")
```

## The Random Forest classifier

Below is the code and the results for the random forest classifier. 

```{r mlTest, include = TRUE, cache = TRUE, echo = TRUE}

library(caret)

# combine datasets, ensure categorical variables are factors
data_ready <- rbind(ci_clean, bf_clean) %>%
  select(-c("hwchazwho")) %>%
  zap_label()

data_ready$urbanhh <- as.factor(data_ready$urbanhh)
data_ready$aglandyn <- as.factor(data_ready$aglandyn)
data_ready$wealthqhh <- as.factor(data_ready$wealthqhh)
data_ready$stunting <- as.factor(data_ready$stunting)

# create index to get training and test samples
train_index <- createDataPartition(data_ready$stunting, p=0.80, list=FALSE)
test_dataset <- data_ready[-train_index,]
train_dataset <- data_ready[train_index,]

set.seed(12)
control <- trainControl(method ="repeatedcv", 
                        repeats = 3, 
                        classProbs = F, 
                        summaryFunction = multiClassSummary)       

rf_haz <- train(stunting ~ ., 
                data = train_dataset, 
                method = "rf", 
                trControl = control, 
                preProcess = c("center", "scale"))

haz_predict <- predict(rf_haz, test_dataset)
confusionMatrix(factor(haz_predict), factor(test_dataset$stunting)) 

```

Overall, the accuracy of the model is not great at around 67%. Additonally, there was a lot of confusion where a lot of 0's were classified as 1's. Therefore, the model is not very good. According to the below graph, the most important variables were many of the environmental variables, like NDVI, precipitation, temperature, and amount of cropland. Sensitivity to environmental variables is consistent with literature exploring these associations between climate and childhood health (Bakhtsiyarava, Grace, and Nawrotzki 2018). These results indicate that more consideration needs to be taken when constructing this model, perhaps taking childhood age in relation to HAZ as a consideration (Alderman and Heady 2018), increasing the sample size, and finding other relevant variables.

```{r var-imp}

varImp(rf_haz) %>% plot()
rf_haz$finalModel

```

Associations can be seen between HAZ and mean temperature, as well as between HAZ and mean NDVI. Maximum temperature extremes seems to have a negative impact on HAZ scores, where higher NDVI values correlates somewhat with positive HAZ scores. 

```{r haz-temp-ndvi}
data_viz <- rbind(ci_clean, bf_clean)

# mean HAZ vs. mean temperature
data_viz %>% 
  filter(urbanhh == 2) %>%
  group_by(ndvi_mean) %>%
  dplyr::summarise(
    mean_haz = mean(hwchazwho),
    cropland = first(cropland),
    temp = first(temp_mean) 
  ) %>%
  ggplot(mapping = aes(x = scale(temp, center = F), y = mean_haz/100)) +
  geom_point(alpha = 2/3) +
  geom_smooth(method = "lm", se = F, color = "darkblue") +
  labs(
    title = "Mean HAZ vs. Mean Maximum Temperature per Cluster",
    subtitle = "For Côte d'Ivoire (2011) and Burkina Faso (2010) combined",
    x = "Mean Max Temperature (scaled)",
    y = "Mean HAZ"
  )

# mean HAZ vs. mean NDVI
data_viz %>% 
  filter(urbanhh == 1) %>%
  group_by(ndvi_mean) %>%
  dplyr::summarise(
    mean_haz = mean(hwchazwho),
    cropland = first(cropland),
    temp = first(temp_mean)
  ) %>%
  ggplot(mapping = aes(x = ndvi_mean, y = mean_haz/100)) +
  geom_point(alpha = 2/3) +
  geom_smooth(method = "lm", se = F, color = "darkblue") +
  labs(
    title = "Mean HAZ vs. Mean NDVI per Cluster",
    subtitle = "For Côte d'Ivoire (2011) and Burkina Faso (2010) combined",
    x = "Mean NDVI",
    y = "Mean HAZ"
  )

```

## Concluding remarks

In this project, I explored factors driving childhood stunting using DHS-IPUMS data to move toward a predictive model of low HAZ. Overall, the Random Forest binary classifier was 67% accurate and contained of misclassified values. Performance may be imporved if HAZ were distinguished by age groups, if environmental data were aggregated differently, if sample size was increased, and if other relevant variables were included. As is consistent with current literature on climate and child nutrition, many of the environmental variables had high levels of importance for the classifier. 

## References

Alderman, Harold, and Derek Headey. 2018. “The Timing of Growth Faltering Has Important Implications for Observational Analyses of the Underlying Determinants of Nutrition Outcomes.” PLOS ONE 13 (4): e0195904. https://doi.org/10.1371/journal.pone.0195904.

Bakhtsiyarava, Maryia, Kathryn Grace, and Raphael J. Nawrotzki. 2017. “Climate, Birth Weight, and Agricultural Livelihoods in Kenya and Mali.” American Journal of Public Health 108 (S2): S144–50. https://doi.org/10.2105/AJPH.2017.304128.

Boyle, Elizabeth Heger, Miriam King and Matthew Sobek. IPUMS-Demographic and Health Surveys: Version 7 [dataset]. Minnesota Population Center and ICF International, 2019. https://doi.org/10.18128/D080.V7

Gubert, Muriel Bauermann, and Maria Helena D’Aquino Benício. 2010. “Use of a Predictive Model for Food Insecurity Estimates in Brazil.” Archivos Latinoamericanos De Nutricion 7.

Onis, Mercedes de, and Francesco Branca. 2016. “Childhood Stunting: A Global Perspective.” Maternal & Child Nutrition 12 (S1): 12–26. https://doi.org/10.1111/mcn.12231.

UNICEF. 2020. Source: https://data.unicef.org/topic/nutrition/malnutrition/
